---
title: "County-Level Socioeconomic Determinants of Health Outcomes"
author: "Maliha Anwar"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    fig-width: 8
    fig-height: 5
---

## Introduction

This analysis examines how county-level socioeconomic conditions are associated with health outcomes across approximately 3,000 U.S. counties. Using data from the [County Health Rankings & Roadmaps](https://www.countyhealthrankings.org/) 2025 release, I wanted explore the relationship between key structural factors such as including income, health insurance coverage, educational attainment, and employment and premature mortality.

**Research Question:** How do county-level socioeconomic factors (median household income, uninsurance rate, educational attainment, and unemployment) predict premature death rates in the United States?

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(broom)
library(gtsummary)
library(broom.helpers)
```

## Data Import

The data comes from the County Health Rankings 2025 CSV Analytic Data file, which contains 796 columns across 3,200+ counties. The raw CSV file contains two header rows: the first provides human-readable variable names, while the second contains standardized variable codes (e.g., `v001_rawvalue`). To ensure consistent variable naming, I skipped the first header row during import and retained the the coded variable names.

```{r}

#| label: import
#| message: false

# Import data, skipping the human-readable header row
# Row 1 = human-readable names, Row 2 = variable codes (we use these)
chr_raw <- read_csv(
  "data-raw/analytic_data2025_v3.csv",
  skip = 1,                          # skip the human-readable header
  na = c("", "NA"),
  show_col_types = FALSE
) |>
  clean_names()

# Quick look at dimensions
cat("Rows:", nrow(chr_raw), "| Columns:", ncol(chr_raw))
```

## Data Cleaning

### Select and Rename Variables

The raw file used coded column names. Here's what each of the selected codes mean:

| Code | Measure                                                    |
|------|------------------------------------------------------------|
| v001 | Premature Death — Years of Potential Life Lost per 100,000 |
| v060 | Diabetes Prevalence (proportion)                           |
| v085 | Uninsured (proportion)                                     |
| v063 | Median Household Income (\$)                               |
| v069 | Some College (proportion)                                  |
| v023 | Unemployment (proportion)                                  |
| v009 | Adult Smoking (proportion)                                 |
| v011 | Adult Obesity (proportion)                                 |

```{r}
#| label: select-rename

health_df <- chr_raw |>
  # Remove state-level summary rows (county code "000")
  filter(countycode != "000") |>
  select(
    fips            = fipscode,
    state           = state,
    county          = county,
    premature_death = v001_rawvalue,
    diabetes_pct    = v060_rawvalue,
    uninsured_pct   = v085_rawvalue,
    median_income   = v063_rawvalue,
    pct_college     = v069_rawvalue,
    pct_unemployed  = v023_rawvalue,
    adult_smoking   = v009_rawvalue,
    adult_obesity   = v011_rawvalue
  ) |>
  mutate(
    across(premature_death:adult_obesity, as.numeric)
  )

glimpse(health_df)
```

### Check Missingness

Before dropping any rows, let's see how much data is missing for each variable.

```{r}
#| label: missingness

health_df |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(
    everything(),
    names_to  = "variable",
    values_to = "n_missing"
  ) |>
  mutate(
    pct_missing = round(n_missing / nrow(health_df) * 100, 1)
  ) |>
  arrange(desc(n_missing))
```

### Handle Missing Data

```{r}
#| label: clean

# Drop rows missing the primary outcome variable
health_clean <- health_df |>
  drop_na(premature_death)

cat(
  "Kept:", nrow(health_clean), "counties\n",
  "Dropped:", nrow(health_df) - nrow(health_clean),
  "counties missing premature death data\n"
)
```

### Convert Proportions to Percentages

Several variables are stored as proportions (0 to 1). Let's convert them to percentages for easier interpretation.

```{r}
#| label: convert-pct

health_clean <- health_clean |>
  mutate(
    diabetes_pct   = diabetes_pct * 100,
    uninsured_pct  = uninsured_pct * 100,
    pct_college    = pct_college * 100,
    pct_unemployed = pct_unemployed * 100,
    adult_smoking  = adult_smoking * 100,
    adult_obesity  = adult_obesity * 100
  )

# Save cleaned data
write_csv(health_clean, "data-processed/health_clean.csv")
```

## Exploratory Data Analysis

### Summary Statistics

```{r}
#| label: summary-stats

health_clean |>
  select(premature_death:adult_obesity) |>
  skim()
```

### Distribution of Premature Death Rate

```{r}
#| label: fig-premature-death-dist
#| fig-cap: "Distribution of premature death rates across U.S. counties"

ggplot(health_clean, aes(x = premature_death)) +
  geom_histogram(bins = 50, fill = "#c0392b", alpha = 0.8) +
  labs(
    title    = "Distribution of Premature Death Rate by County",
    subtitle = "Years of Potential Life Lost per 100,000 population",
    x = "YPLL per 100,000",
    y = "Number of Counties"
  ) +
  theme_minimal(base_size = 14)
```

### Income vs. Premature Death

```{r}
#| label: fig-income-scatter
#| fig-cap: "Relationship between median household income and premature death rate"

ggplot(health_clean, aes(x = median_income, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#1a7a6d", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title    = "Higher-Income Counties Tend to Have Lower Premature Death Rates",
    subtitle = "Each point represents one U.S. county",
    x = "Median Household Income",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Uninsured Rate vs. Premature Death

```{r}
#| label: fig-uninsured-scatter
#| fig-cap: "Relationship between uninsured rate and premature death rate"

ggplot(health_clean, aes(x = uninsured_pct, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#2c3e50", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  labs(
    title    = "Counties with Higher Uninsured Rates Tend to Have Higher Premature Death",
    x = "Uninsured (%)",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Correlation Matrix

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of all numeric health and socioeconomic variables"
#| fig-height: 6
#| fig-width: 7

health_clean |>
  select(premature_death:adult_obesity) |>
  cor(use = "pairwise.complete.obs") |>
  corrplot(
    method   = "color",
    type     = "upper",
    tl.col   = "black",
    tl.cex   = 0.85,
    addCoef.col = "black",
    number.cex  = 0.7
  )
```

## Regression Modeling

### Fit the Model

```{r}
#| label: model-fit

model <- lm(
  premature_death ~ median_income + uninsured_pct +
    pct_college + pct_unemployed + diabetes_pct,
  data = health_clean
)

glance(model)
```

### Regression Table

```{r}
#| label: tbl-regression
#| tbl-cap: "Multiple linear regression: predictors of premature death rate"

model |>
  tbl_regression(
    label = list(
      median_income  ~ "Median Household Income ($)",
      uninsured_pct  ~ "Uninsured (%)",
      pct_college    ~ "Some College (%)",
      pct_unemployed ~ "Unemployment (%)",
      diabetes_pct   ~ "Diabetes Prevalence (%)"
    )
  ) |>
  bold_p()
```

### Diagnostic Plots

These plots help us check the assumptions of linear regression: linearity, normality of residuals, and constant variance.

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-height: 8
#| fig-width: 8
#| layout-ncol: 2

# Residuals vs Fitted
aug <- augment(model)

ggplot(aug, aes(.fitted, .resid)) +
  geom_point(alpha = 0.25, color = "#1a7a6d") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#c0392b") +
  labs(
    title = "Residuals vs Fitted Values",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal(base_size = 13)

# Q-Q Plot
ggplot(aug, aes(sample = .resid)) +
  stat_qq(alpha = 0.25, color = "#1a7a6d") +
  stat_qq_line(color = "#c0392b") +
  labs(
    title = "Normal Q-Q Plot of Residuals",
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 13)
```

### Interpret the Results

```{r}
#| label: interpret

# Tidy coefficients with confidence intervals
tidy(model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~round(., 4)))
```

> **Write your interpretation here.** For each predictor, describe:
>
> -   The direction of the relationship (positive or negative)
> -   The magnitude (what does a 1-unit change in the predictor mean for YPLL?)
> -   Statistical significance (check the p-value and confidence interval)
>
> For example: *"For every \$1,000 increase in median household income, premature death rate decreased by approximately X YPLL per 100,000, holding all other variables constant (p \< 0.001)."*

## Limitations

-   **Ecological fallacy:** These are county-level associations. We cannot conclude that individuals with lower incomes have higher premature death rates — only that counties with lower median incomes tend to have higher premature death rates.
-   **Cross-sectional design:** This is a snapshot from one time period. We cannot establish causation.
-   **Missing data:** Some counties were excluded due to missing outcome data, which may introduce bias.
-   **Unmeasured confounders:** Factors like access to healthcare facilities, environmental exposures, and racial/ethnic composition are not included in this model.

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
