---
title: "County-Level Socioeconomic Determinants of Health Outcomes"
author: "Maliha Anwar"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    fig-width: 8
    fig-height: 5
---

## Introduction

Premature death is measured as Years of Potential Life Lost (YPLL) before age 75 and varies widely across U.S. counties. This analysis explores how key structural factors such as income, health insurance coverage, educational attainment, and employment are associated with premature mortality. Understanding these relationships is beneficial in informing public health policy and resource allocation by identifying structural conditions linked to higher risk of early death.

**Research Question:** How do county-level socioeconomic factors predict premature death rates in the United States?

This is a county-level (ecological) analysis, therefore the pattern describe counties, not individuals. I explore this question using the 2025 [County Health Rankings & Roadmaps](https://www.countyhealthrankings.org/) data through exploratory visualization, correlation analysis, and multiple linear regression.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(broom)
library(gtsummary)
```

## Data Import

The County Health Rankings CSV has two header rows: the first with human-readable names ("Premature Death raw value") and the second with variable codes ("v001_rawvalue"). To ensure consistent variable naming, I skipped the first header row and used the coded variable names. They are also are easier to work with programmatically.

The file contains 796 columns across 3,200+ counties.

```{r}
#| label: import
#| message: false

chr_raw <- read_csv(
  "data-raw/analytic_data2025_v3.csv",
  skip = 1,
  na = c("", "NA"),
  show_col_types = FALSE
) |>
  clean_names()

cat("Rows:", nrow(chr_raw), "| Columns:", ncol(chr_raw))
```

## Data Cleaning

### Selecting Variables

Out of 796 columns, I selected eight measures that are relevant to the research question. The table below maps the coded column names to what they actually represent.

| Code | Measure                 | Units            |
|------|-------------------------|------------------|
| v001 | Premature death         | YPLL per 100,000 |
| v060 | Diabetes prevalence     | Proportion       |
| v085 | Uninsured               | Proportion       |
| v063 | Median household income | Dollars          |
| v069 | Some college            | Proportion       |
| v023 | Unemployment            | Proportion       |
| v009 | Adult smoking           | Proportion       |
| v011 | Adult obesity           | Proportion       |

```{r}
#| label: select-rename

health_df <- chr_raw |>
  # Row with countycode "000" is the state-level summary — remove these
  filter(countycode != "000") |>
  select(
    fips            = fipscode,
    state           = state,
    county          = county,
    premature_death = v001_rawvalue,
    diabetes_pct    = v060_rawvalue,
    uninsured_pct   = v085_rawvalue,
    median_income   = v063_rawvalue,
    pct_college     = v069_rawvalue,
    pct_unemployed  = v023_rawvalue,
    adult_smoking   = v009_rawvalue,
    adult_obesity   = v011_rawvalue
  ) |>
  mutate(
    across(premature_death:adult_obesity, as.numeric)
  )

cat("Counties selected:", nrow(health_df))
```

### Checking for Missing Data

Before dropping anything from the dataset, I wanted to see which variables have missing values and how many. This helps decide whether to drop rows, impute, or accept the gaps.

```{r}
#| label: missingness

missing_summary <- health_df |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(
    everything(),
    names_to  = "variable",
    values_to = "n_missing"
  ) |>
  mutate(pct_missing = round(n_missing / nrow(health_df) * 100, 1)) |>
  arrange(desc(n_missing))

missing_summary
```

### Cleaning Decisions

Based on the missingness (see above), here's what I decided to do:

-   **Drop counties missing premature death** (the outcome variable), since there is no way to analyze the outcome if it's not there.
-   **Keep counties with missing predictors for now.** R's `lm()` handles this with complete-case analysis by default, which means it will only use rows where all variables in the model are present. For the first round of analysis, I decided not to impute.
-   **Convert proportions to percentages.** Several variables (diabetes, uninsured, smoking, etc.) were stored as proportions between 0 and 1. I find that converting them to percentages makes the regression coefficients easier to interpret. For example: A "1 percentage-point increase" is more intuitive than "a 0.01 increase."

```{r}
#| label: clean

health_clean <- health_df |>
  drop_na(premature_death) |>
  mutate(
    diabetes_pct   = diabetes_pct * 100,
    uninsured_pct  = uninsured_pct * 100,
    pct_college    = pct_college * 100,
    pct_unemployed = pct_unemployed * 100,
    adult_smoking  = adult_smoking * 100,
    adult_obesity  = adult_obesity * 100
  )

cat(
  "Counties retained:", nrow(health_clean), "\n",
  "Counties dropped:", nrow(health_df) - nrow(health_clean),
  "(missing premature death data)\n"
)

# Save for use in other scripts or the Shiny dashboard
write_csv(health_clean, "data-processed/health_clean.csv")
```

## Exploratory Data Analysis

### Summary Statistics

I started the analysis by taking a broad look at all the numeric variables, their ranges, means, and how spread out they are.

```{r}
#| label: summary-stats

health_clean |>
  select(premature_death:adult_obesity) |>
  skim()
```

### Distribution of Premature Death Rates

Before modeling, I wanted to see what the outcome variable looks like. Is it roughly normal? Are there any extreme outliers?

```{r}
#| label: fig-outcome-dist
#| fig-cap: "Distribution of premature death rates across U.S. counties, 2025"

ggplot(health_clean, aes(x = premature_death)) +
  geom_histogram(bins = 50, fill = "#c0392b", alpha = 0.8) +
  labs(
    title = "Distribution of Premature Death Rate by County",
    subtitle = "Years of Potential Life Lost per 100,000",
    x = "YPLL per 100,000",
    y = "Number of Counties"
  ) +
  theme_minimal(base_size = 14)
```

### Income vs. Premature Death

This is the relationship I'm most interested in. Income is often considered a "fundamental cause" of health in public health literature. In the U.S., income shapes access to almost everything ranging from housing, food, healthcare to safe environments/neighborhoods.

```{r}
#| label: fig-income-scatter
#| fig-cap: "Counties with higher median incomes tend to have lower premature death rates"

ggplot(health_clean, aes(x = median_income, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#1a7a6d", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Median Household Income vs. Premature Death Rate",
    subtitle = "Each point is one U.S. county",
    x = "Median Household Income",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Uninsured Rate vs. Premature Death

Insurance coverage is another factor to consider. Counties with more uninsured residents may have less access to preventive care and treatment.

```{r}
#| label: fig-uninsured-scatter
#| fig-cap: "Counties with higher uninsured rates tend to have higher premature death rates"

ggplot(health_clean, aes(x = uninsured_pct, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#2c3e50", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  labs(
    title = "Uninsured Rate vs. Premature Death Rate",
    subtitle = "Each point is one U.S. county",
    x = "Uninsured (%)",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Correlation Matrix

Before putting variables into a regression together, I want to check how strongly they correlate with each other. If two predictors are highly correlated (roughly r \> 0.7), including both could cause multicollinearity. It means the predictors are so closely related that the model has trouble separating which one is actually driving the effect.

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of health and socioeconomic variables"
#| fig-height: 6
#| fig-width: 7

cor_matrix <- health_clean |>
  select(premature_death:adult_obesity) |>
  cor(use = "pairwise.complete.obs")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.cex      = 0.85,
  addCoef.col = "black",
  number.cex  = 0.7
)
```

## Regression Modeling

### The Model

I used multiple linear regression to see how each socioeconomic factor relates to premature death rate when other factors are accounted for.

The model: `premature_death ~ median_income + uninsured_pct + pct_college + pct_unemployed + diabetes_pct`

I included diabetes prevalence because it captures some of the chronic disease burden in a county. I left smoking and obesity out because they are highly correlated with the other predictors and with each other, which would cause multicollinearity issue.

```{r}
#| label: model-fit

model <- lm(
  premature_death ~ median_income + uninsured_pct +
    pct_college + pct_unemployed + diabetes_pct,
  data = health_clean
)

# Model summary: R-squared, F-statistic, etc.
glance(model) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, nobs)
```

### Coefficient Table

This table shows the estimated effect of each predictor on premature death rate, along with 95% confidence intervals and p-values. Statistically significant predictors (p \< 0.05) are bolded.

```{r}
#| label: tbl-regression
#| tbl-cap: "Multiple linear regression: predictors of county-level premature death rate"

model |>
  tbl_regression(
    label = list(
      median_income  ~ "Median Household Income ($)",
      uninsured_pct  ~ "Uninsured (%)",
      pct_college    ~ "Some College (%)",
      pct_unemployed ~ "Unemployment (%)",
      diabetes_pct   ~ "Diabetes Prevalence (%)"
    )
  ) |>
  bold_p()

```

### Checking Model Assumptions

Linear regression relies on a couple of key assumptions here:
1. The residuals (the differences between predicted and actual values) are roughly normally distributed
2. Their spread stays consistent across the range of predictions. 
These two plots help check that.

**What to look for:**

-   **Residuals vs. Fitted:** If the model fits well, the points should look randomly scattered around the zero line with no obvious pattern. If they fan out in a funnel shape, that's a sign the model's predictions are less reliable at certain values.
-   **Q-Q Plot:** The points should roughly follow the diagonal line. If they curve away at the ends, it means the residuals aren't as normally distributed as the model assumes.

```{r}
#| label: fig-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-height: 4.5
#| layout-ncol: 2

aug <- augment(model)

# Residuals vs Fitted
ggplot(aug, aes(.fitted, .resid)) +
  geom_point(alpha = 0.25, color = "#1a7a6d") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#c0392b") +
  labs(
    title = "Residuals vs. Fitted Values",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal(base_size = 13)

# Q-Q plot
ggplot(aug, aes(sample = .resid)) +
  stat_qq(alpha = 0.25, color = "#1a7a6d") +
  stat_qq_line(color = "#c0392b") +
  labs(
    title = "Normal Q-Q Plot of Residuals",
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 13)
```

### Interpreting the Results
The model explains about 53% of the variation in premature death rates across counties (adjusted R² = 0.525). In other words, roughly half of the differences between counties come down to these five factors. The other half is likely driven by things not in this model, like racial segregation, environmental conditions, and healthcare access.

All five predictors were statistically significant (p < 0.001).

**Median Household Income** (β = -0.08): For every additional dollar in median income, premature death drops by 0.08 YPLL per 100,000. That sounds small, but it adds up. $10,000 difference in income translates to about 800 fewer YPLL. The scatter plot also shows this isn't a straight-line relationship: premature death drops sharply between $25,000 and $75,000, then levels off.

**Diabetes Prevalence** (β = 677): This was the strongest predictor. A 1 percentage-point increase in diabetes prevalence is linked to 677 additional YPLL per 100,000. Diabetes typically shows up as a comorbidity with heart disease and hypertension, so this coefficient is likely reflecting a broader pattern of chronic illness.

**Unemployment (β = 183)**: A 1 percentage-point increase in unemployment is linked to 183 additional YPLL per 100,000. Unemployment impacts many things that affect health such as income, insurance, mental health, and access to resources.

**Uninsured**(β = -44): I found this one surprising. A higher uninsured rate was actually associated with lower premature death, which is the opposite of what I expected. Looking at the scatter plot, the relationship between uninsured rate and premature death is weak and mostly flat. One explanation I could come up with is that once income and the other predictors are accounted for, the uninsured variable may be picking up something else, for example, counties with high uninsured rates in the South and Southwest tend to have younger populations, which could bring YPLL down despite poor coverage. This was a good reminder that regression coefficients shouldn't be read in isolation.

**Some College** (β = -25): A 1 percentage-point increase in college attendance is linked to 25 fewer YPLL per 100,000. In public health, it's well established that education levels are correlated with higher life expectancy. Education connects to health in several ways like health literacy, better job opportunities and income, and being able to navigate the healthcare system.

## Limitations
There are several important caveats to keep in mind as you go through this analysis:

-   **Ecological fallacy.** This analysis uses county-level data. I can say that counties with lower incomes tend to have higher premature death rates, but I cannot make the same claim for individuals based on this analysis alone.
-   **Cross-sectional design.** This is a single snapshot in time, not a trend. I can describe associations but not establish causation. Lower income doesn't necessarily *cause* higher premature death. There may be shared underlying causes driving both.
-   **Missing data.** 64 counties were dropped due to missing premature death data. These tend to be very small or rural counties, so the results may not fully represent the most underserved areas.
-   **Unmeasured confounders.** There are many factors that influence health outcomes are not in this model — racial segregation, environmental exposures, healthcare facility density, housing quality to name a few. The R-squared value reflects how much variation the chosen predictors explain, but there's clearly more to the story.
-   **Model specification.** I used a linear model, which assumes a straight-line relationship between each predictor and the outcome. The scatter plots suggest this is a reasonable, but the relationship between income and premature death is likely nonlinear (the loess curve hints at this).

## Session Info

```{r}
#| label: session-info

sessionInfo()
```

