---
title: "County-Level Socioeconomic Determinants of Health Outcomes"
author: "Maliha Anwar"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    fig-width: 8
    fig-height: 5
---

## Introduction

Premature death is measured as Years of Potential Life Lost (YPLL) before age 75 and varies widely across U.S. counties. This analysis explores how key structural factors such as income, health insurance coverage, educational attainment, and employment are associated with premature mortality. Understanding these relationships is beneficial in informing public health policy and resource allocation by identifying structural conditions linked to higher risk of early death.

**Research Question:** How do county-level socioeconomic factors predict premature death rates in the United States?

This is a county-level (ecological) analysis, therefore the pattern describe counties, not individuals. I explore this question using the 2025 [County Health Rankings & Roadmaps](https://www.countyhealthrankings.org/) data through exploratory visualization, correlation analysis, and multiple linear regression.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(skimr)
library(corrplot)
library(broom)
library(gtsummary)
```

## Data Import

The County Health Rankings CSV has two header rows: the first with human-readable names ("Premature Death raw value") and the second with variable codes ("v001_rawvalue"). To ensure consistent variable naming, I skipped the first header row and used the coded variable names. They are also are easier to work with programmatically.

The file contains 796 columns across 3,200+ counties.

```{r}
#| label: import
#| message: false

chr_raw <- read_csv(
  "data-raw/analytic_data2025_v3.csv",
  skip = 1,
  na = c("", "NA"),
  show_col_types = FALSE
) |>
  clean_names()

cat("Rows:", nrow(chr_raw), "| Columns:", ncol(chr_raw))
```

## Data Cleaning

### Selecting Variables

Out of 796 columns, I selected eight measures that are relevant to the research question. The table below maps the coded column names to what they actually represent.

| Code | Measure                 | Units            |
|------|-------------------------|------------------|
| v001 | Premature death         | YPLL per 100,000 |
| v060 | Diabetes prevalence     | Proportion       |
| v085 | Uninsured               | Proportion       |
| v063 | Median household income | Dollars          |
| v069 | Some college            | Proportion       |
| v023 | Unemployment            | Proportion       |
| v009 | Adult smoking           | Proportion       |
| v011 | Adult obesity           | Proportion       |

```{r}
#| label: select-rename

health_df <- chr_raw |>
  # Row with countycode "000" is the state-level summary — remove these
  filter(countycode != "000") |>
  select(
    fips            = fipscode,
    state           = state,
    county          = county,
    premature_death = v001_rawvalue,
    diabetes_pct    = v060_rawvalue,
    uninsured_pct   = v085_rawvalue,
    median_income   = v063_rawvalue,
    pct_college     = v069_rawvalue,
    pct_unemployed  = v023_rawvalue,
    adult_smoking   = v009_rawvalue,
    adult_obesity   = v011_rawvalue
  ) |>
  mutate(
    across(premature_death:adult_obesity, as.numeric)
  )

cat("Counties selected:", nrow(health_df))
```

### Checking for Missing Data

Before dropping anything from the dataset, I wanted to see which variables have missing values and how many. This helps decide whether to drop rows, impute, or accept the gaps.

```{r}
#| label: missingness

missing_summary <- health_df |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  pivot_longer(
    everything(),
    names_to  = "variable",
    values_to = "n_missing"
  ) |>
  mutate(pct_missing = round(n_missing / nrow(health_df) * 100, 1)) |>
  arrange(desc(n_missing))

missing_summary
```

### Cleaning Decisions

Based on the missingness (see above), here's what I decided to do:

-   **Drop counties missing premature death** (the outcome variable), since there is no way to analyze the outcome if it's not there.
-   **Keep counties with missing predictors for now.** R's `lm()` handles this with complete-case analysis by default, which means it will only use rows where all variables in the model are present. For the first round of analysis, I decided not to impute.
-   **Convert proportions to percentages.** Several variables (diabetes, uninsured, smoking, etc.) were stored as proportions between 0 and 1. I find that converting them to percentages makes the regression coefficients easier to interpret. For example: A "1 percentage-point increase" is more intuitive than "a 0.01 increase."

```{r}
#| label: clean

health_clean <- health_df |>
  drop_na(premature_death) |>
  mutate(
    diabetes_pct   = diabetes_pct * 100,
    uninsured_pct  = uninsured_pct * 100,
    pct_college    = pct_college * 100,
    pct_unemployed = pct_unemployed * 100,
    adult_smoking  = adult_smoking * 100,
    adult_obesity  = adult_obesity * 100
  )

cat(
  "Counties retained:", nrow(health_clean), "\n",
  "Counties dropped:", nrow(health_df) - nrow(health_clean),
  "(missing premature death data)\n"
)

# Save for use in other scripts or the Shiny dashboard
write_csv(health_clean, "data-processed/health_clean.csv")
```

## Exploratory Data Analysis

### Summary Statistics

I started the analysis by taking a broad look at all the numeric variables, their ranges, means, and how spread out they are.

```{r}
#| label: summary-stats

health_clean |>
  select(premature_death:adult_obesity) |>
  skim()
```

### Distribution of Premature Death Rates

Before modeling, I wanted to see what the outcome variable looks like. Is it roughly normal? Are there any extreme outliers?

```{r}
#| label: fig-outcome-dist
#| fig-cap: "Distribution of premature death rates across U.S. counties, 2025"

ggplot(health_clean, aes(x = premature_death)) +
  geom_histogram(bins = 50, fill = "#c0392b", alpha = 0.8) +
  labs(
    title = "Distribution of Premature Death Rate by County",
    subtitle = "Years of Potential Life Lost per 100,000",
    x = "YPLL per 100,000",
    y = "Number of Counties"
  ) +
  theme_minimal(base_size = 14)
```

### Income vs. Premature Death

This is the relationship I'm most interested in. Income is often considered a "fundamental cause" of health in public health literature. In the U.S., income shapes access to almost everything ranging from housing, food, healthcare to safe environments/neighborhoods.

```{r}
#| label: fig-income-scatter
#| fig-cap: "Counties with higher median incomes tend to have lower premature death rates"

ggplot(health_clean, aes(x = median_income, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#1a7a6d", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Median Household Income vs. Premature Death Rate",
    subtitle = "Each point is one U.S. county",
    x = "Median Household Income",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Uninsured Rate vs. Premature Death

Insurance coverage is another factor to consider. Counties with more uninsured residents may have less access to preventive care and treatment.

```{r}
#| label: fig-uninsured-scatter
#| fig-cap: "Counties with higher uninsured rates tend to have higher premature death rates"

ggplot(health_clean, aes(x = uninsured_pct, y = premature_death)) +
  geom_point(alpha = 0.25, color = "#2c3e50", size = 1.5) +
  geom_smooth(method = "loess", color = "#c0392b", se = TRUE, alpha = 0.15) +
  labs(
    title = "Uninsured Rate vs. Premature Death Rate",
    subtitle = "Each point is one U.S. county",
    x = "Uninsured (%)",
    y = "YPLL per 100,000"
  ) +
  theme_minimal(base_size = 14)
```

### Correlation Matrix

Before putting variables into a regression together, I want to check how strongly they correlate with each other. If two predictors are highly correlated (roughly r \> 0.7), including both could cause multicollinearity. It means the predictors are so closely related that the model has trouble separating which one is actually driving the effect.

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation matrix of health and socioeconomic variables"
#| fig-height: 6
#| fig-width: 7

cor_matrix <- health_clean |>
  select(premature_death:adult_obesity) |>
  cor(use = "pairwise.complete.obs")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  tl.col      = "black",
  tl.cex      = 0.85,
  addCoef.col = "black",
  number.cex  = 0.7
)
```

## Regression Modeling

### The Model

I used multiple linear regression to see how each socioeconomic factor relates to premature death rate when other factors are accounted for.

The model: `premature_death ~ median_income + uninsured_pct + pct_college + pct_unemployed + diabetes_pct`

I included diabetes prevalence because it captures some of the chronic disease burden in a county. I left smoking and obesity out because they are highly correlated with the other predictors and with each other, which would cause multicollinearity issue.

```{r}
#| label: model-fit

model <- lm(
  premature_death ~ median_income + uninsured_pct +
    pct_college + pct_unemployed + diabetes_pct,
  data = health_clean
)

# Model summary: R-squared, F-statistic, etc.
glance(model) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, nobs)
```

### Coefficient Table

This table shows the estimated effect of each predictor on premature death rate, along with 95% confidence intervals and p-values. Statistically significant predictors (p \< 0.05) are bolded.

```{r}
#| label: tbl-regression
#| tbl-cap: "Multiple linear regression: predictors of county-level premature death rate"

model |>
  tbl_regression(
    label = list(
      median_income  ~ "Median Household Income ($)",
      uninsured_pct  ~ "Uninsured (%)",
      pct_college    ~ "Some College (%)",
      pct_unemployed ~ "Unemployment (%)",
      diabetes_pct   ~ "Diabetes Prevalence (%)"
    )
  ) |>
  bold_p()

```

### 
